A semantics of core AUTOSAR

### 2015-08-22 paper/paper.tex
Some LaTeX formatting

\newcommand{\V}[1]{\mbox{\it #1}}
\newcommand{\Code}{\V{Code}}
\newcommand{\Cont}{\V{Cont}}
\newcommand{\State}{\V{State}}
\newcommand{\Act}{\V{Act}}
\newcommand{\Upd}{\V{Upd}}
\newcommand{\Xs}{\V{Xs}}
\newcommand{\Vs}{\V{Vs}}
\newcommand{\Cs}{\V{Cs}}

\newcommand{\void}{\mbox{\bf void}}
\newcommand{\nil}{\mbox{\bf nil}}
\newcommand{\true}{\mbox{\bf true}}
\newcommand{\false}{\mbox{\bf false}}
\newcommand{\idle}{\mbox{\bf idle}}
\newcommand{\pending}{\mbox{\bf pending}}
\newcommand{\serving}[2]{\mbox{\bf serving}(#1,#2)}

\newcommand{\adr}[1]{\mbox{\em #1}}
\newcommand{\sep}{\hspace{2ex}|\hspace{2ex}}

\newcommand{\runnable}[4]{\mbox{\bf runnable}(#1,#2,#3,#4)}
\newcommand{\rinst}[4]{\mbox{\bf rinst}(#1,#2,#3,#4)}
\newcommand{\excl}[2]{\mbox{\bf excl}(#1,#2)}
\newcommand{\irv}[2]{\mbox{\bf irv}(#1,#2)}
\newcommand{\qelem}[2]{\mbox{\bf qelem}(#1,#2)}
\newcommand{\delem}[3]{\mbox{\bf delem}(#1,#2,#3)}
\newcommand{\opres}[2]{\mbox{\bf opres}(#1,#2)}
\newcommand{\timer}[3]{\mbox{\bf timer}(#1,#2, #3)}

\newcommand{\Enter}[2]{\mbox{rte\_Enter}(#1,#2)}
\newcommand{\Exit}[2]{\mbox{rte\_Exit}(#1,#2)}
\newcommand{\IrvWrite}[3]{\mbox{rte\_IrvWrite}(#1,#2,#3)}
\newcommand{\IrvRead}[2]{\mbox{rte\_IrvRead}(#1,#2)}
\newcommand{\Send}[4]{\mbox{rte\_Send}(#1,#2,#3,#4)}
\newcommand{\Receive}[3]{\mbox{rte\_Receive}(#1,#2,#3)}
\newcommand{\Write}[4]{\mbox{rte\_Write}(#1,#2,#3,#4)}
\newcommand{\Read}[3]{\mbox{rte\_Read}(#1,#2,#3)}
\newcommand{\Invalidate}[3]{\mbox{rte\_Invalidate}(#1,#2,#3)}
\newcommand{\IsUpdated}[3]{\mbox{rte\_IsUpdated}(#1,#2,#3)}
\newcommand{\Call}[4]{\mbox{rte\_Call}(#1,#2,#3,#4)}
\newcommand{\Result}[3]{\mbox{rte\_Result}(#1,#2,#3)}
\newcommand{\Return}[1]{\mbox{return}(#1)}

@@

\newcommand{\ent}{\mbox{\small ent}}
\newcommand{\ex}{\mbox{\small ex}}

@@

An AUTOSAR model is primarily a structure of interconnected {\bf software components} (SWCs). Links between SWCs are called {\bf connectors}, which attach to {\bf ports} exposed by the SWCs. A port is either {\bf required} or {\bf provided}, and then classified according to what {\bf interface} it exposes: {\bf sender-receiver}, {\bf client-server}, {\bf trigger} and {\bf mode-switch} are common interface types. A sender-receiver interface is an aggregation of one or more {\bf data-elements}, with each element characterized by the type of data it carries and optional further communication specification details. A client-server interface likewise consists of one or more {\bf operations}, that each give the signature of a procedure call and the possible errors that may result.

Some components merely act as containers of other components structures, these are called {\bf composition} SWCs. In contrast, {\bf atomic} SWCs define their own behavior in terms of {\bf runnables} and {\bf inter-runnable variables}, primarily. A runnable denotes a sequential and (normally) terminating piece of code, although its actual behavioral definition is typically relegated to external C or Matlab files rather than being part of the AUTOSAR model proper. Instead, an AUTOSAR runnable  only provides constraints on what ports and inter-runnable variables the actual implementation can access; as well as {\bf event} declarations that determine under what conditions it will get executed.

A runnable under execution is called a {\bf runnable instance}, and these may execute concurrently even if they belong to the same SWC. Each runnable indicates whether it may be instantiated concurrently with itself. Runnable code can also use {\bf exclusive areas} (a form of mutex semaphores) to further control concurrent behavior.

To interact with ports, inter-runnable variables and exclusive areas, runnables utilize what is abstractly known as the {\bf virtual function bus} (VFB). Concretely, the VFB takes the shape of a C programming interface to the {\bf run-time environment} (RTE), custom-generated for each runnable on basis of its access constraints. Behind the RTE, a collection of OS kernels, communication stacks and other basic software modules implement the VBF services for the platform at hand. Complete AUTOSAR designs also typically contain {\bf ECU} and {\bf task assignments}, which are manually built tables that control how SWCs and runnables map onto the available hardware and OS resources, respectively.

@@

Components and ports do not carry any dynamic state besides the state of their constituent parts, so they need no explicit representation as process terms. However, in order to still be able to talk about model elements using their locally scoped names, we employ a hierarchical addressing scheme, such that address $\adr{R:I} $ means runnable $R$ of component $I$ and $\adr{E:P:I}$ means element $E$ of port $P$ within component $I$. The following meta-variable conventions will be used:
$$
\begin{array}{rcll}
	I	& \in		& \mbox{Component names} \\
	R	& \in		& \mbox{Runnable names} \\
	U	& \in		& \mbox{Port names} \\
	E	& \in		& \mbox{Sender-receiver element names} \\
	O	& \in		& \mbox{Client-server operation names} \\
	S	& \in		& \mbox{Inter-runnable variable names} \\
	X	& \in		& \mbox{Exclusive area names} \\
	\\
	A	& \in		& \mbox{Addresses} \\
	A	& ::=		& \adr{R:I}   \sep   \adr{E:P:I}   \sep   \adr{O:U:I}   \sep   \adr{S:I}   \sep   \adr{X:I}   \\
\end{array}
$$
Since AUTOSAR components can be arbitrarily nested, the component names $I$ can also be considered to have a similarly nested internal structure (component $I_1$ within component $I_2$ within component $I_3$, etc). We can fully ignore this detail here, though, as the hierarchical layout of components in an AUTOSAR system has no run-time implications \cite{}.

@@

In the untyped setting of this report, a value $V$ can be any data item computed and communicated by an AUTOSAR system, including the unit value $\void$ and the error codes that might be returned from RTE calls. Meta-variables $N$, $B$ and $T$ range over natural numbers, boolean values and (floating point) time values, respectively. The client $C$ of a runnable instance is the address of the invoking port operation (if the runnable was invoked by a client-server call), or $\void$ otherwise. An activation state $\Act$ either toggles between $\idle$ and $\pending$, or it has the special shape $\serving{\Cs}{\Vs}$ (in the case of a runnable triggered by client-server calls). Formally,
$$
\begin{array}{rcll}
	V	& \in		& \mbox{Values} \\
	N	& \in		& \mbox{Natural numbers} \\
	B	& \in		& \mbox{Boolean values} \\
	T	& \in		& \mbox{Time values} \\
	C	& ::=		& \adr{O:U:I} \sep \void \\
	\Act	& ::=		& \idle \sep \pending \sep \serving{\Cs}{\Vs} \\
\end{array}
$$
We write $\Vs$ for a sequence of values, $\Cs$ for a sequence of clients, etc. Sequences may be empty, which is written $\epsilon$. Left (right) concatenation of an element with a sequence is written $\adr{V:Vs}$ ($\adr{Vs:V}$).

@@

Our semantic approach defines the meaning of an AUTOSAR system as the possible chains of state changes that can be applied to the system's initial state; or equivalently, as the set of {\em traces} that can be generated from the corresponding process term $P_0$. A trace is a possibly infinite sequence of transitions
$$
\begin{array}{c}
	P_0 \red{L_1} P_1 \red{L_2} P_2 \red{L_3} \cdots
\end{array}
$$
where each step
$$
\begin{array}{c}
	P_{i-1} \red{L_i} P_i
\end{array}
$$
states that the system state captured as process term $P_{i-1}$ can evolve into state $P_i$ by means of a single transition labelled $L_i$.

The label $L$ of a transition can be ow two kinds, indicating that the affected process term either "says" or "hears" something during the transition. A label also always consists of an address $A$ paired with some additional detail $D$, as captured in the following label grammar:
$$
\begin{array}{rcll}
	L	& ::=		& A!D		& \hspace{4em}\mbox{Say $A$ and $D$} \\
		&		& A?D		& \hspace{4em}\mbox{Hear $A$ and $D$} \\
\end{array}
$$
The possible forms of the extra label information $D$ will be introduced as the different transition rules are defined.

@@

The effect of these agreement rules closely resembles the idea of a {\em broadcast}: what one process says may be heard by every other process in a large parallel composition. This behavior serves us well, because in general, what one AUTOSAR runnable does may have impact on many -- if not all -- parts of the executing system. It is however important not to confuse this broadcasting notion with any particular form of AUTOSAR communication. The way transition labels distribute over a parallel system is merely a technical aspect of our process calculus, and will be used to express several different AUTOSAR communication semantics, among other things.

@@

\item For each runnable $R$ of each component $I$, a term
$$
\begin{array}{c}
	\runnable{\adr{R:I}}{0.0}{\Act}{0}
\end{array}
$$
where $\Act$ is $\serving{\epsilon}{\epsilon}$ if $R$ is triggered by an {\em OperationInvokedEvent}; else $\Act$ is $\pending$ if $R$ is triggered by an {\em InitEvent}; otherwise $\Act$ is $\idle$.
% SWS_Rte_03524 (SWS_Rte_03526?) forbids mixing OperationInvokedEvents with other events for the same runnable

@@

\item For each data element $E$ of each required port $U$ of each atomic component $I$, a term
$$
\begin{array}{c}
	\qelem{\adr{E:U:I}}{\epsilon}
\end{array}
$$
if the {\em swImplPolicy} attribute of element $E$ is {\em queued}; otherwise, a term
$$
\begin{array}{c}
	\delem{\adr{E:U:I}}{\false}{\mbox{NO\_DATA}}
\end{array}
$$

@@

\item For each {\em TimingEvent} triggering each runnable $R$ of each atomic component $I$, a term
$$
\begin{array}{c}
	\timer{\adr{R:I}}{T}{0.0}
\end{array}
$$
where $T$ is the period of the event.

@@

Some further static model information will also be referenced by the transition rules. We refer to the {\em connectors} of a model using a binary relation $\Rightarrow$ on addresses, defined as
$$
	\adr{E:$U_1$:$I_1$} \Rightarrow \adr{E:$U_2$:$I_2$} \;\;\mbox{for all elements $E$ of port $U_2$}
$$
if there is a connector in the model from provided sender-receiver port $U_1$ of component $I_1$ to required port $U_2$ of component $I_2$; and
$$
	\adr{O:$U_1$:$I_1$} \Rightarrow \adr{O:$U_2$:$I_2$} \;\;\mbox{for all operations $O$ of port $U_1$}
$$
if there is a connector in the model from required client-server port $U_1$ of component $I_1$ to provided port $U_2$ of component $I_2$.

We expect the {\em implementation} of a runnable to be available as a continuation $\Cont$, even though the real AUTOSAR model will be refering to concrete functions in external C/Matlab files. We will further abstract away from actual numbers of input and output parameters to these function by assuming that records (structs) are used as the single continuation argument and result whenever the arity so requires. Absent arguments and results will be replaced by the unit value $\void$.

>>>>

\newcommand{\V}[1]{\text{\it #1}}
\newcommand{\Code} {\V{Code}}
\newcommand{\Cont} {\V{Cont}}
\newcommand{\State}{\V{State}}
\newcommand{\Act}  {\V{Act}}
\newcommand{\Upd}  {\V{Upd}}
\newcommand{\Xs}   {\V{Xs}}
\newcommand{\Vs}   {\V{Vs}}
\newcommand{\Cs}   {\V{Cs}}

\newcommand{\kw}[1]{\textbf{#1}}
\newcommand{\void}    {\kw{void}}
\newcommand{\nil}     {\kw{nil}}
\newcommand{\true}    {\kw{true}}
\newcommand{\false}   {\kw{false}}
\newcommand{\idle}    {\kw{idle}}
\newcommand{\pending} {\kw{pending}}
\newcommand{\serving}[2]{\kw{serving}(#1,#2)}

\newcommand{\adr}[1]{\emph{#1}}
\newcommand{\sep}{\hspace{2ex}|\hspace{2ex}}

\newcommand{\runnable}[4]{\kw{runnable}(#1,#2,#3,#4)}
\newcommand{\rinst}   [4]{\kw{rinst}(#1,#2,#3,#4)}
\newcommand{\excl}    [2]{\kw{excl}(#1,#2)}
\newcommand{\irv}     [2]{\kw{irv}(#1,#2)}
\newcommand{\qelem}   [2]{\kw{qelem}(#1,#2)}
\newcommand{\delem}   [3]{\kw{delem}(#1,#2,#3)}
\newcommand{\opres}   [2]{\kw{opres}(#1,#2)}
\newcommand{\timer}   [3]{\kw{timer}(#1,#2, #3)}

\newcommand{\rte}[1]{\text{rte\_#1}}
\newcommand{\Enter}     [2]{\rte{Enter}(#1,#2)}
\newcommand{\Exit}      [2]{\rte{Exit}(#1,#2)}
\newcommand{\IrvWrite}  [3]{\rte{IrvWrite}(#1,#2,#3)}
\newcommand{\IrvRead}   [2]{\rte{IrvRead}(#1,#2)}
\newcommand{\Send}      [4]{\rte{Send}(#1,#2,#3,#4)}
\newcommand{\Receive}   [3]{\rte{Receive}(#1,#2,#3)}
\newcommand{\Write}     [4]{\rte{Write}(#1,#2,#3,#4)}
\newcommand{\Read}      [3]{\rte{Read}(#1,#2,#3)}
\newcommand{\Invalidate}[3]{\rte{Invalidate}(#1,#2,#3)}
\newcommand{\IsUpdated} [3]{\rte{IsUpdated}(#1,#2,#3)}
\newcommand{\Call}      [4]{\rte{Call}(#1,#2,#3,#4)}
\newcommand{\Result}    [3]{\rte{Result}(#1,#2,#3)}
\newcommand{\Return}    [1]{\text{return}(#1)}

@@

\newcommand{\ent}{\text{\small ent}}
\newcommand{\ex}{\text{\small ex}}

@@

An AUTOSAR model is primarily a structure of interconnected \textbf{software components} (SWCs). Links between SWCs are called \textbf{connectors}, which attach to \textbf{ports} exposed by the SWCs. A port is either \textbf{required} or \textbf{provided}, and then classified according to what \textbf{interface} it exposes: \textbf{sender-receiver}, \textbf{client-server}, \textbf{trigger} and \textbf{mode-switch} are common interface types. A sender-receiver interface is an aggregation of one or more \textbf{data-elements}, with each element characterized by the type of data it carries and optional further communication specification details. A client-server interface likewise consists of one or more \textbf{operations}, that each give the signature of a procedure call and the possible errors that may result.

Some components merely act as containers of other components structures, these are called \textbf{composition} SWCs. In contrast, \textbf{atomic} SWCs define their own behavior in terms of \textbf{runnables} and \textbf{inter-runnable variables}, primarily. A runnable denotes a sequential and (normally) terminating piece of code, although its actual behavioral definition is typically relegated to external C or Matlab files rather than being part of the AUTOSAR model proper. Instead, an AUTOSAR runnable  only provides constraints on what ports and inter-runnable variables the actual implementation can access; as well as \textbf{event} declarations that determine under what conditions it will get executed.

A runnable under execution is called a \textbf{runnable instance}, and these may execute concurrently even if they belong to the same SWC. Each runnable indicates whether it may be instantiated concurrently with itself. Runnable code can also use \textbf{exclusive areas} (a form of mutex semaphores) to further control concurrent behavior.

To interact with ports, inter-runnable variables and exclusive areas, runnables utilize what is abstractly known as the \textbf{virtual function bus} (VFB). Concretely, the VFB takes the shape of a C programming interface to the \textbf{run-time environment} (RTE), custom-generated for each runnable on basis of its access constraints. Behind the RTE, a collection of OS kernels, communication stacks and other basic software modules implement the VBF services for the platform at hand. Complete AUTOSAR designs also typically contain \textbf{ECU} and \textbf{task assignments}, which are manually built tables that control how SWCs and runnables map onto the available hardware and OS resources, respectively.

@@

Components and ports do not carry any dynamic state besides the state of their constituent parts, so they need no explicit representation as process terms. However, in order to still be able to talk about model elements using their locally scoped names, we employ a hierarchical addressing scheme, such that address $\adr{R:I}$ means runnable $R$ of component $I$ and $\adr{E:P:I}$ means element $E$ of port $P$ within component $I$. The following meta-variable conventions will be used:
$$
\begin{array}{rcll}
	I	& \in		& \text{Component names} \\
	R	& \in		& \text{Runnable names} \\
	U	& \in		& \text{Port names} \\
	E	& \in		& \text{Sender-receiver element names} \\
	O	& \in		& \text{Client-server operation names} \\
	S	& \in		& \text{Inter-runnable variable names} \\
	X	& \in		& \text{Exclusive area names} \\
	\\
	A	& \in		& \text{Addresses} \\
	A	& ::=		& \adr{R:I}   \sep   \adr{E:P:I}   \sep   \adr{O:U:I}   \sep   \adr{S:I}   \sep   \adr{X:I}   \\
\end{array}
$$
Since AUTOSAR components can be arbitrarily nested, the component names $I$ can also be considered to have a similarly nested internal structure (component $I_1$ within component $I_2$ within component $I_3$, etc). We can fully ignore this detail here, though, as the hierarchical layout of components in an AUTOSAR system has no run-time implications \cite{}.

@@

In the untyped setting of this report, a value $V$ can be any data item computed and communicated by an AUTOSAR system, including the unit value $\void$ and the error codes that might be returned from RTE calls. Meta-variables $N$, $B$ and $T$ range over natural numbers, boolean values and (floating point) time values, respectively. The client $C$ of a runnable instance is the address of the invoking port operation (if the runnable was invoked by a client-server call), or $\void$ otherwise. An activation state $\Act$ either toggles between $\idle$ and $\pending$, or it has the special shape $\serving{\Cs}{\Vs}$ (in the case of a runnable triggered by client-server calls). Formally,
$$
\begin{array}{rcll}
	V	& \in		& \text{Values} \\
	N	& \in		& \text{Natural numbers} \\
	B	& \in		& \text{Boolean values} \\
	T	& \in		& \text{Time values} \\
	C	& ::=		& \adr{O:U:I} \sep \void \\
	\Act	& ::=		& \idle \sep \pending \sep \serving{\Cs}{\Vs} \\
\end{array}
$$
We write $\Vs$ for a sequence of values, $\Cs$ for a sequence of clients, etc. Sequences may be empty, which is written $\epsilon$. Left (right) concatenation of an element with a sequence is written $\adr{V:Vs}$ ($\adr{Vs:V}$).

@@

Our semantic approach defines the meaning of an AUTOSAR system as the possible chains of state changes that can be applied to the system's initial state; or equivalently, as the set of \emph{traces} that can be generated from the corresponding process term $P_0$. A trace is a possibly infinite sequence of transitions
$$
\begin{array}{c}
	P_0 \red{L_1} P_1 \red{L_2} P_2 \red{L_3} \cdots
\end{array}
$$
where each step
$$
\begin{array}{c}
	P_{i-1} \red{L_i} P_i
\end{array}
$$
states that the system state captured as process term $P_{i-1}$ can evolve into state $P_i$ by means of a single transition labelled $L_i$.

The label $L$ of a transition can be ow two kinds, indicating that the affected process term either "says" or "hears" something during the transition. A label also always consists of an address $A$ paired with some additional detail $D$, as captured in the following label grammar:
$$
\begin{array}{rcll}
	L	& ::=		& A!D		& \hspace{4em}\text{Say $A$ and $D$} \\
		&		& A?D		& \hspace{4em}\text{Hear $A$ and $D$} \\
\end{array}
$$
The possible forms of the extra label information $D$ will be introduced as the different transition rules are defined.

@@

The effect of these agreement rules closely resembles the idea of a \emph{broadcast}: what one process says may be heard by every other process in a large parallel composition. This behavior serves us well, because in general, what one AUTOSAR runnable does may have impact on many -- if not all -- parts of the executing system. It is however important not to confuse this broadcasting notion with any particular form of AUTOSAR communication. The way transition labels distribute over a parallel system is merely a technical aspect of our process calculus, and will be used to express several different AUTOSAR communication semantics, among other things.

@@

\item For each runnable $R$ of each component $I$, a term
$$
\begin{array}{c}
	\runnable{\adr{R:I}}{0.0}{\Act}{0}
\end{array}
$$
where $\Act$ is $\serving{\epsilon}{\epsilon}$ if $R$ is triggered by an \emph{OperationInvokedEvent}; else $\Act$ is $\pending$ if $R$ is triggered by an \emph{InitEvent}; otherwise $\Act$ is $\idle$.
% SWS_Rte_03524 (SWS_Rte_03526?) forbids mixing OperationInvokedEvents with other events for the same runnable

@@

\item For each data element $E$ of each required port $U$ of each atomic component $I$, a term
$$
\begin{array}{c}
	\qelem{\adr{E:U:I}}{\epsilon}
\end{array}
$$
if the \emph{swImplPolicy} attribute of element $E$ is \emph{queued}; otherwise, a term
$$
\begin{array}{c}
	\delem{\adr{E:U:I}}{\false}{\text{NO\_DATA}}
\end{array}
$$

@@

\item For each \emph{TimingEvent} triggering each runnable $R$ of each atomic component $I$, a term
$$
\begin{array}{c}
	\timer{\adr{R:I}}{T}{0.0}
\end{array}
$$
where $T$ is the period of the event.

@@

Some further static model information will also be referenced by the transition rules. We refer to the \emph{connectors} of a model using a binary relation $\Rightarrow$ on addresses, defined as
$$
	\adr{E:$U_1$:$I_1$} \Rightarrow \adr{E:$U_2$:$I_2$} \;\;\text{for all elements $E$ of port $U_2$}
$$
if there is a connector in the model from provided sender-receiver port $U_1$ of component $I_1$ to required port $U_2$ of component $I_2$; and
$$
	\adr{O:$U_1$:$I_1$} \Rightarrow \adr{O:$U_2$:$I_2$} \;\;\text{for all operations $O$ of port $U_1$}
$$
if there is a connector in the model from required client-server port $U_1$ of component $I_1$ to provided port $U_2$ of component $I_2$.

We expect the \emph{implementation} of a runnable to be available as a continuation $\Cont$, even though the real AUTOSAR model will be refering to concrete functions in external C/Matlab files. We will further abstract away from actual numbers of input and output parameters to these function by assuming that records (structs) are used as the single continuation argument and result whenever the arity so requires. Absent arguments and results will be replaced by the unit value $\void$.
